Layer based routing to obtain quality optimization of video streaming in the Future Internet

### 2012-12-06 paper/paper.tex
Paper

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Layer based routing to obtain quality optimization of scalable video streaming in the Future Internet}

@@

\begin{abstract}
%\boldmath
In TCP/IP networks, there is no inherent support for Quality of Service.
Routing is done on the minimum hop count from sender to receiver, and all traffic is best-effort traffic. 
This is unsatisfactory for applications such as video streaming. 
This paper presents an OpenFlow controller application which provides 
Quality of Service (\textit{QoS}) support for Scalable Video Streaming (\textit{SVC}) in an OpenFlow network. 
First we route the base layer of SVC encoded videos as a 
high priority QoS flow while the enhancement layers are routed at a lower QoS or as a best-effort flow. 
Second, we provide failure recovery so the OpenFlow network can converge back to an optimal 
state when links fail. We show that using different costs to calculate paths for different flows 
can improve the overall throughput of video streams when network congestion and delay occur.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

\section{Introduction}
In a TCP/IP-based network, each router or switch contains a flow table containing the 
next hop for each destination, based on the shortest path calculation. 
Media applications, such as video streaming, 
require no packet loss and minimal network delay due to their high Quality of Service (QoS) dependency. 
When these QoS flows are sent over the same path as other best effort traffic they 
are subject to network congestion and delay which renders an unsatisfactory outcome. 
Several efforts have been made to support QoS in a TCP/IP-based network, such as DiffServ and IntServ, but these 
solutions only work well in managed networks.
Scalable Video Coding (SVC) encodes the stream in a base layer along with several enhancement layers. 
In order to obtain correct output it is important that the base layer has no 
packet loss and minimal delay, which is not possible without support for QoS. 
To solve this issue we need the routing algorithms to detect QoS flows and route these flows differently than 
normal best effort traffic.
To achieve this goal in video streaming, we route the base layer of SVC encoded video at a high QoS 
while the enhancement layers are routed either at a lower QoS or as a best effort flow. 
We also provide failure recovery so the network 
can converge back to an optimal state when links go down.\\

We use OpenFlow in order to provide QoS support for scalable video streaming. 
OpenFlow is a protocol for switches, allowing to control the packet forwarding process.
This way, networks are not only configurable, but also programmable.
Each OpenFlow Switch has a flow table, which is used for packet lookup and 
forwarding, and a secure channel to an external controller (see Figure \ref{fig:openflow1}). 
The flow table consists of flow entries.
Each flow table entry matches a traffic pattern, by matching the header fields.
The header fields that can be matched in OpenFlow 1.0 can be seen in Table \ref{tab:matchedflows}.
The controller makes the network programmable by managing the switch over 
the secure channel using the OpenFlow protocol.
The controller installs new entries in the flow table to manage the network traffic.
Entries have a hard and a soft timeout. A soft timeout means the entry gets removed if it has not been used for a certain time.
A hard timeout means the flow will be removed after the timeout, even if it is still in use.
There are many implementations of controllers available for OpenFlow switches.
The implementation of the controller we will use is Floodlight, because it is open source and widely used.

@@

\section{Related work}
In our research we identified a need for separate forwarding actions per flow.
The same principal has been discussed in research papers from Princeton University. 
In their research the aim was to apply load-balancing during service discovery of web services 
when all requests had to be processed by a central component. 
A problem occurs when each request is forwarded to a centralized component responsible 
for service discovery, which leads to network congestion. By using OpenFlow, the researchers at 
Princeton University managed to set separate forwarding actions for client requests which match 
certain criteria. These actions allow forwarding entities to redirect client requests to an 
appropriate service instance without interference from the centralized component. Even though 
the use case may be different, the principal is the same as the one applied in our research. 
When both best effort and QoS flows follow the same paths, chances for congestion and packet loss 
to occur increase significantly. OpenFlow allows us to set different paths for both flows, which 
improves the throughput of our QoS flows and also decrease the network load on links used by best 
effort traffic.\\

The idea to provide a minimal quality for SVC encoded videos originates from 
research described in [4] and [5]. By routing the base layer as a lossless-QoS flow we can 
guarantee a minimal quality to the user. Ideally, the enhancement layers will also be delivered 
without packet loss or delay but when congestion occurs we demote the enhancement layers to best 
effort traffic and prioritize the base layer. Research discussed in [4] also attempts to route 
the enhancement layers as a lossy-QoS flow. Our research goes further and also enables fast failure 
recovery when a link fails.\\

\section{Architecture}
Each OpenFlow network contains at least one controller which is responsible for managing all the switches. 
A default OpenFlow switch forwards packets 
to the controller if the packet does not match any forwarding entry. 
We route SVC encoded video streams as a QoS flow, where we consider the base layer as a 
high priority QoS flow and the enhancement layer as either a lower QoS flow or as best effort traffic. 
These principals are shown in Figure \ref{fig_sim}. A filter on the controller separates QoS flows from 
other traffic, after which the controller can choose a path for the different layers. 
When the controller calculates a path for a layer, the new forwarding rules are sent to the switches.\\

\begin{figure*}[htb]
\centering
\includegraphics[width=\linewidth]{fig/SVC_OpenFlow_flow}
\caption{Simulation Results}
\label{fig_sim}
\end{figure*}

We defined a cost function which can be used to calculate the cost of a path.
The cost function is based on the following observations. (Insert ref here)
\begin{itemize}
	\item When high priority QoS flows are forwarded on the same link as best effort traffic, there will be packet loss in the best effort flow to prioritize the QoS flow.
	\item Metrics for QoS flows must be different than those used for best effort traffic. The shortest path and QoS path must be different to obtain good results. 
	If both follow the same path, the best effort traffic will experience more packet loss.
	\item Calculating a path for a new QoS flow does not change the best effort routing tables. The packet loss of best effort traffic caused by QoS flows must be considered during path selection.
	\item The length of alternative paths for QoS flows should still be kept as short as possible to optimize transfer time.
	\item Best effort traffic can be used as a reference to retrieve metrics from a switch or link, such as packet loss or to estimate the amount of traffic on that route.
\end{itemize}

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,IEEEexample}

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Layer based routing to obtain quality optimization of video streaming in the Future Internet}

@@

\begin{abstract}
%\boldmath
In TCP/IP networks, there is no inherent support for Quality of Service.
Routing is done on the minimum hop count from sender to receiver, and all traffic is best-effort traffic. 
This is unsatisfactory for applications such as video streaming. 
This paper presents an OpenFlow controller application which provides 
Quality of Service (\textit{QoS}) support for video streaming using Scalable Video Coding (\textit{SVC}) in an OpenFlow network. 
First we route the base layer of SVC encoded videos as a 
high priority QoS flow while the enhancement layers are routed at a lower QoS or as a best-effort flow. 
Second, we provide failure recovery so the OpenFlow network can converge back to an optimal 
state when links fail. We show that using different costs to calculate paths for different flows 
can improve the overall throughput of video streams when network congestion and delay occur.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

\section{Introduction}
In a TCP/IP-based network, each router or switch contains a flow table containing the 
next hop for each destination, based on the shortest path calculation. 
Media applications, such as video streaming, 
require no packet loss and minimal network delay. 
When these video flows are sent over the same path as other best effort traffic they 
are subject to network congestion and delay which renders an unsatisfactory outcome. 
Several efforts have been made to give precedence to certain types of traffic, defined as Quality of Service (QoS).
There are solutions available for QoS support in a TCP/IP-based network, 
such as DiffServ \cite{diffserv} and IntServ \cite{intserv}, but these 
solutions only work well in managed networks.
Scalable Video Coding (SVC) \cite{schwarz2007overview} encodes the stream in a base layer along with several enhancement layers. 
In order to obtain qualitative video streams, it is important that the base layer has no 
packet loss and minimal delay, which is not possible without support for QoS. 
To solve this issue we need the routing algorithms to detect QoS flows and route these flows differently than 
normal best effort traffic.
To achieve this goal in video streaming, we route the base layer of SVC encoded video at a high QoS 
while the enhancement layers are routed either at a lower QoS or as a best effort flow. 
We also provide failure recovery so the network 
can converge back to an optimal state when links fail.\\

We use OpenFlow \cite{mckeown2008openflow} in order to provide QoS support for scalable video streaming. 
OpenFlow is a protocol for switches, allowing to control the packet forwarding process.
This way, networks are not only configurable, but also programmable.
Each OpenFlow Switch has a flow table, which is used for packet lookup and 
forwarding, and a secure channel to an external controller (see Figure \ref{fig:openflow1}). 
The flow table consists of flow entries.
Each flow table entry matches a traffic pattern, by matching the header fields.
The header fields that can be matched in OpenFlow 1.0 can be seen in Table \ref{tab:matchedflows}.
The controller makes the network programmable by managing the switch over 
the secure channel using the OpenFlow protocol.
The controller installs new entries in the flow table to manage the network traffic.
Entries have a hard and a soft timeout. A soft timeout means the entry gets removed if it has not been used for a certain time.
A hard timeout means the flow will be removed after the timeout, even if it is still in use.
There are many implementations of controllers available for OpenFlow switches.
The implementation of the controller we will use is Floodlight \cite{floodlight}, because it is open source and widely used.

@@

\section{Related work}
In our research we identified a need for separate forwarding actions per flow.
Wang, R. et al. \cite{wang2011openflow} follow the same principal.
The aim was to apply load-balancing during service discovery of web services 
when all requests had to be processed by a central component.
A problem occurs when each request is forwarded to a centralized component responsible 
for service discovery, which leads to network congestion. By using OpenFlow, the researchers at 
Princeton University managed to set separate forwarding actions for client requests which match 
certain criteria. These actions allow forwarding entities to redirect client requests to an 
appropriate service instance without interference from the centralized component. Even though 
the use case may be different, the principal is the same as the one applied in our research. 
When both best effort and QoS flows follow the same path, chances of congestion and packet loss 
occurring increase significantly. OpenFlow allows us to set different paths for both flows, which 
improves the throughput of our QoS flows and also decrease the network load on links used by best 
effort traffic.\\

The idea to provide a minimal quality for SVC encoded videos originates from 
research described in \cite{egilmez2011scalable} and \cite{civanlar2010qos}. 
By routing the base layer as a lossless-QoS flow we can 
guarantee a minimal quality to the user. Ideally, the enhancement layers will also be delivered 
without packet loss or delay but when congestion occurs we demote the enhancement layers to best 
effort traffic and prioritize the base layer. Research discussed in \cite{egilmez2011scalable} 
also attempts to route the enhancement layers as a lossy-QoS flow. 
Our research goes further and also enables fast failure 
recovery when a link fails.\\

\section{Design considerations}
Each OpenFlow network contains at least one controller which is responsible for managing all the switches. 
A default OpenFlow switch forwards packets 
to the controller if the packet does not match any forwarding entry. 
We route SVC encoded video streams as a QoS flow, where we consider the base layer as a 
high priority QoS flow and the enhancement layer as either a lower QoS flow or as best effort traffic. 
These principals are shown in Figure \ref{fig_sim}. A filter on the controller separates QoS flows from 
other traffic, after which the controller can choose a path for the different layers. 
When the controller calculates a path for a layer, the new forwarding rules are sent to the switches.\\

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{fig/SVC_OpenFlow_flow}
\caption{Simulation Results}
\label{fig_sim}
\end{figure}

We defined a cost function which can be used to calculate the cost of a path.
The cost function is based on the following observations.
\begin{itemize}
	\item When high priority QoS flows are forwarded on the same link as best effort traffic, there will be packet loss in the best effort flow to prioritize the QoS flow.
	\item Metrics for QoS flows must be different than those used for best effort traffic. The shortest path and QoS path must be different to obtain good results. 
	If both follow the same path, the best effort traffic will experience more packet loss.
	\item Calculating a path for a new QoS flow does not change the best effort routing tables. The packet loss of best effort traffic caused by QoS flows must be considered during path selection.
	\item The length of alternative paths for QoS flows should still be kept as short as possible to optimize transfer time.
	\item Best effort traffic can be used as a reference to retrieve metrics from a switch or link, such as packet loss or to estimate the amount of traffic on that route.
\end{itemize}

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
\bibliography{paper}
